# Annotate Metabolites from Feature Tables {#sec-annotate-metabolites}

> *"You have successfully processed 100 gigabytes of raw LC-MS data. Your statistical analysis reveals 42 features that are significantly upregulated in your disease cohort. But a feature is just a mathematical coordinate: $m/z$ 205.097 at 4.2 minutes. What molecule is it? Until you answer that question, your biological narrative is stalled."*

Metabolite annotation is the primary bottleneck in untargeted metabolomics. This chapter teaches you how to transition from anonymous $m/z$ features to putative chemical identities using the R ecosystem.

::: {.callout-warning title="The One Mistake to Avoid"}
Reporting a database match as an identification. A matching mass is a *hypothesis* (MSI Level 3), not a named compound. Reserve the word "identified" for annotations confirmed by MS/MS or an authentic standard.
:::

## Learning Objectives

By the end of this chapter you will be able to:

- Explain the MSI confidence levels (1–4) and the annotation funnel
- Deconvolve isotopes and adducts from a feature table with CAMERA
- Compute exact monoisotopic masses and adduct $m/z$ values with `MetaboCoreUtils`
- Search observed features against an exact-mass reference database within an instrument tolerance
- Assemble a metabolite annotation report that distinguishes putative identities from unknowns

## The Annotation Funnel

Translating an exact mass to a biological identity is rarely a 1:1 mapping. A single metabolite can produce dozens of distinct features in a mass spectrometer due to isotopes, adduct formation (e.g., binding with sodium or potassium), and in-source fragmentation. Conversely, a single exact mass might correspond to hundreds of different isomeric structures (e.g., glucose and fructose have the exact same mass).

To solve this, annotation is structured as a funnel of increasing confidence:

```{mermaid}
flowchart TD
    A[Feature Matrix<br/>Thousands of m/z & RT pairs] -->|Deconvolution<br/>CAMERA| B[Pseudo-Spectra<br/>Grouped Adducts & Isotopes]
    B -->|Exact Mass Search<br/>MetaboCoreUtils| C[Putative Candidates<br/>Database Matching]
    C -->|Spectral Matching<br/>MsCoreUtils| D[Library Annotations<br/>MS/MS Scoring]
    D -->|Authentic Standards| E[Confirmed Metabolites]
    
    style A fill:#D7E6FB,stroke:#27408B
    style B fill:#FBE0FA,stroke:#B000B0
    style C fill:#D7E6FB,stroke:#27408B
    style D fill:#FBE0FA,stroke:#B000B0
    style E fill:#ffffff,stroke:#27408B,stroke-width:2px
```

### The MSI Confidence Levels

Before assigning a name to a feature, you must understand the **Metabolomics Standards Initiative (MSI)** confidence levels. Overstating confidence is a common trap in the literature.

- **Level 1 (Confirmed):** The feature's retention time, exact mass, and MS/MS spectrum identically match an authentic chemical standard run on the *exact same instrument*.
- **Level 2 (Putative Annotation):** The feature matches the exact mass and MS/MS spectrum of a public library (e.g., MassBank, MoNA), but no authentic standard was run locally.
- **Level 3 (Putative Class):** The spectral evidence suggests a chemical class (e.g., "a phosphatidylcholine") but the exact isomeric structure cannot be determined.
- **Level 4 (Unknown):** The exact mass is known, but it cannot be matched to any database or spectrum.

This chapter focuses on achieving **Level 2 and Level 3** annotations computationally.

---

## Step 1: Deconvoluting Features with CAMERA

If glucose ($\text{C}_6\text{H}_{12}\text{O}_6$) elutes from your chromatography column at 2.5 minutes, the mass spectrometer does not just see one peak. It typically sees:

- `[M+H]+`: The protonated molecule
- `[M+Na]+`: The sodium adduct
- `[M+K]+`: The potassium adduct
- `[M+H]+` (13C): The naturally occurring Carbon-13 isotope
- `[M+H-H2O]+`: An in-source water loss fragment

If you run statistical tests directly on the `xcms` feature table, glucose will appear as 5 separate "significant" features. This artificially inflates your results. The `CAMERA` package (Collection of Algorithms for MEtabolite pRofile Annotation) solves this by grouping these related features into a single "pseudo-spectrum" based on their perfectly correlated retention times and peak shapes.

### Implementing CAMERA

To use `CAMERA`, we take the `xcmsSet` or `XCMSnExp` object generated after peak detection and alignment (as covered in Chapter 7).

```{r}
#| message: false
#| warning: false
library(xcms)
library(CAMERA)

# Assume 'xdata' is your fully aligned and grouped XCMSnExp object
# Step 1: Convert to xcmsSet (CAMERA's preferred format)
# xset <- as(xdata, "xcmsSet")

# Step 2: Initialize the CAMERA annotate object
# xa <- xsAnnotate(xset)

# Step 3: Group features by retention time and peak shape correlation
# xa <- groupFWHM(xa, perfwhm = 0.6)
# xa <- groupCorr(xa, calcIso = TRUE, calcCiS = TRUE, calcCaS = TRUE)

# Step 4: Identify Isotopes and Adducts
# xa <- findIsotopes(xa, mzabs = 0.015)
# xa <- findAdducts(xa, polarity = "positive")

# Step 5: Extract the deconvoluted peak table
# annotated_peaks <- getPeaklist(xa)
```

The resulting `annotated_peaks` table will contain new columns:
- `pcgroup`: The pseudo-spectrum group ID. All features with the same `pcgroup` likely originate from the same parent metabolite.
- `isotopes`: Identifies if the feature is an [M+1] or [M+2] isotope.
- `adduct`: Putative adduct annotations (e.g., `[M+Na]+`).

By filtering this table to only keep the primary monoisotopic peaks (often the `[M+H]+` or `[M-H]-` adducts), you massively reduce the dimensionality of your data and focus only on true biological entities.

---

## Step 2: Chemical Math with MetaboCoreUtils

Once you have a deconvoluted list of monoisotopic $m/z$ values, you need to calculate their theoretical neutral masses to search chemical databases.

The `MetaboCoreUtils` package provides high-performance functions for chemical formula and adduct math. It bridges the gap between chemistry and data science.

### Calculating Exact Masses and Adduct $m/z$

Let's calculate the theoretical exact mass of Glucose and see what $m/z$ values we should expect to see in positive ionization mode.

```{r}
library(MetaboCoreUtils)

# 1. Calculate the exact monoisotopic mass of Glucose
glucose_formula <- "C6H12O6"
glucose_mass <- calculateMass(glucose_formula)

cat("Exact Mass of Glucose:", glucose_mass, "Da\n")

# 2. Define the adducts we expect to see in Positive Mode ESI
expected_adducts <- c("[M+H]+", "[M+Na]+", "[M+K]+", "[M+NH4]+")

# 3. Calculate the expected m/z values for these adducts
glucose_mz <- mass2mz(glucose_mass, adduct = expected_adducts)

print(glucose_mz)
```

### Working Backwards: From $m/z$ to Mass

In untargeted metabolomics, we work in reverse. We observe an $m/z$ in the instrument, and we must infer the neutral mass to search databases like HMDB or KEGG.

```{r}
# We observe a significant feature at m/z 203.0526
observed_mz <- 203.0526

# If we assume this feature is a Sodium adduct [M+Na]+
# What is the neutral mass of the original molecule?
neutral_mass <- mz2mass(observed_mz, adduct = "[M+Na]+")

cat("Inferred Neutral Mass:", neutral_mass, "Da\n")
```

Notice that the inferred neutral mass (`180.0634`) matches the exact mass of Glucose. 

::: {.callout-tip}
#### Adduct Assumption Risk
If you assume a feature is `[M+H]+` but it is actually `[M+Na]+`, your calculated neutral mass will be off by roughly ~22 Da. You will search the database for the completely wrong molecule. This highlights why `CAMERA` deconvolution (which clusters adducts together to define the true neutral mass) is critical *before* database searching.
:::

---

## Step 3: Exact Mass Database Searching

Armed with estimated neutral masses (or `[M+H]+` $m/z$ values), we can query biological databases. 

For this example, we will simulate a local database lookup using `dplyr`. In practice, packages like `CompoundDb` allow you to build sophisticated SQLite databases from HMDB or MassBank, but the mathematical logic of the join remains the same.

### Handling parts-per-million (ppm) Error

Mass spectrometers are not perfectly accurate. An Orbitrap might have 3 ppm error, while a Q-TOF might have 10 ppm error. We cannot use a strict `==` operator to match our observed mass to the database mass. We must use a range join based on the instrument's ppm tolerance.

```{r}
library(dplyr)
library(tidyr)

# 1. Simulate a tiny reference database
reference_db <- tibble::tribble(
  ~compound_name, ~formula,   ~exact_mass,
  "Glucose",      "C6H12O6",  180.063388,
  "Fructose",     "C6H12O6",  180.063388, # Isomer of Glucose!
  "Caffeine",     "C11H14N4O2", 204.101452,
  "Serotonin",    "C11H12N2O",  176.094963,
  "Alanine",      "C3H7NO2",  89.047678
)

# 2. Simulate our significant features from XCMS/CAMERA
# We assume we have already deduced these are [M+H]+ adducts, 
# so we converted them back to neutral exact masses.
my_features <- tibble::tribble(
  ~feature_id, ~observed_mass, ~retention_time,
  "FT001",     180.0638,       2.5,
  "FT002",     204.1011,       5.8,
  "FT003",     300.1234,       8.2   # Unknown mass
)
```

Now we perform the matching algorithm. We calculate the allowable lower and upper mass bounds for each observed feature based on a 5 ppm tolerance, and then filter the database for any compounds that fall within that window.

```{r}
# Define instrument tolerance
ppm_tolerance <- 5

# Calculate the mass window for our observed features
my_features <- my_features %>%
  mutate(
    mass_lower = observed_mass - (observed_mass * ppm_tolerance / 1e6),
    mass_upper = observed_mass + (observed_mass * ppm_tolerance / 1e6)
  )

# Perform a non-equi join (Range Join) using cross_join and filter
annotated_features <- my_features %>%
  cross_join(reference_db) %>%
  filter(exact_mass >= mass_lower & exact_mass <= mass_upper) %>%
  mutate(
    # Calculate the actual ppm error of the match
    ppm_error = abs(observed_mass - exact_mass) / exact_mass * 1e6
  ) %>%
  select(feature_id, observed_mass, compound_name, formula, ppm_error)

print(annotated_features)
```

### The Isomer Problem

Look closely at the output for `FT001`. It matched *both* Glucose and Fructose. This is biologically accurate—a mass spectrometer measuring exact mass alone cannot distinguish between stereoisomers. 

This highlights why exact mass matching only achieves **Level 3 (Putative Class)** or weak **Level 2** confidence. To distinguish Glucose from Fructose, you must either:
1. Match their distinct fragmentation patterns (MS/MS spectral library search).
2. Compare their retention times against authentic standards run on your exact chromatography setup.

---

## Step 4: Synthesizing the Final Reporting Table

Once you have mapped your features to putative IDs, you must fold these annotations back into your quantitative feature matrix for downstream statistical modeling and pathway analysis.

Because one feature can match multiple isomers, we usually collapse the multiple matches into a single, delimited string so the feature matrix retains its strict $Features \times Samples$ rectangular dimensions.

```{r}
# Collapse multiple isomer matches into a single string
collapsed_annotations <- annotated_features %>%
  group_by(feature_id, observed_mass) %>%
  summarize(
    putative_identities = paste(compound_name, collapse = " | "),
    formulas = paste(unique(formula), collapse = " | "),
    min_ppm_error = min(ppm_error),
    .groups = "drop"
  )

# Re-join with the original feature list to see what remained unknown
final_report <- my_features %>%
  select(feature_id, observed_mass, retention_time) %>%
  left_join(collapsed_annotations, by = c("feature_id", "observed_mass")) %>%
  mutate(
    annotation_status = ifelse(is.na(putative_identities), 
                               "Level 4 (Unknown)", 
                               "Level 3 (Putative Mass Match)")
  )

print(final_report)
```

## Summary

Metabolite annotation is an iterative process of shrinking the universe of chemical possibilities:

1. **CAMERA** shrinks thousands of redundant features down to representative monoisotopic neutral masses by recognizing adducts and isotopes.
2. **MetaboCoreUtils** provides the strict chemical math required to interconvert masses and $m/z$ ratios based on ionization rules.
3. **PPM Range Joins** map your experimental observations to massive structural databases (like HMDB) while accounting for instrument error.
4. **MSI Confidence Levels** govern how you report these findings, preventing you from overstating isomer-blind exact mass matches.

In the next chapter, we will take these Level 3 exact-mass annotations and upgrade them to Level 2 by matching their MS/MS fragmentation patterns against spectral libraries.

## Exercises

1. **Adduct Math:** Using `MetaboCoreUtils`, calculate the exact mass of Caffeine (`C11H14N4O2`). What $m/z$ would you expect to see for its `[M+Na]+` adduct? What about a `[M-H]-` adduct in negative mode?
2. **Tolerance Effects:** In the `dplyr` exact mass search code above, change the `ppm_tolerance` from 5 to 50 (mimicking an older TOF instrument). How does this affect the number of potential matches and your confidence in them?
3. **Annotation Summaries:** Write a brief `dplyr` pipeline using the `final_report` table to calculate the percentage of features that were successfully annotated versus those that remain "Level 4 (Unknown)".

## Session Information

```{r}
sessionInfo()
```