# Build and Audit a Quantitative Proteomics Data Object

You cannot measure a protein directly. You measure peptides, then argue your way up to the protein — propagating every filtering decision and every missing value as you go. Get that bookkeeping wrong and a "differentially abundant protein" can be an artifact of which peptides you happened to keep. This chapter builds the `QFeatures` object that keeps the PSM → peptide → protein hierarchy honest, linked, and auditable.

::: {.callout-warning title="The One Mistake to Avoid"}
Aggregating peptides to proteins *before* filtering. Summarizing over contaminants, one-hit-wonders, and mostly-missing peptides bakes noise straight into the protein value. Filter first, then aggregate.
:::

## Learning Objectives

By the end of this chapter you will be able to:

- Describe how `QFeatures` organises the PSM → peptide → protein hierarchy
- Import quantitative proteomics data and construct a `QFeatures` object
- Filter, log-transform, normalize, and aggregate features across levels
- Diagnose data quality with PCA and missing-value visualisation
- Run a basic differential test and compare standard versus robust aggregation

## Understanding Quantitative MS Data

### Quantitation Methodologies

There are several approaches to quantitative proteomics, each with distinct advantages:

```{r}
library(QFeatures)
library(tidyverse)
library(limma)
library(ggplot2)
library(pheatmap)

factoextra_available <- suppressWarnings(requireNamespace("factoextra", quietly = TRUE))
```

#### Label-free MS1: Extracted Ion Chromatograms (XIC)

In label-free quantitation, precursor peaks matching identified peptides are integrated over retention time.

#### Labelled MS2: Isobaric Tagging (TMT/iTRAQ)

Isobaric tags allow multiplexed quantitation where peptides from different samples are chemically labeled and analyzed together.

#### Label-free MS2: Spectral Counting

Simple counting of peptide-spectrum matches assigned to each protein.

#### Labelled MS1: SILAC

Stable isotope labeling allows direct comparison between heavy and light labeled samples.

## The QFeatures Framework

### QFeatures Class Structure

`QFeatures` extends the `MultiAssayExperiment` class to handle the hierarchical nature of MS data (spectra → peptides → proteins).

```{mermaid}
%%| fig-width: 10
%%| fig-height: 6
flowchart TD
    subgraph QF["QFeatures Object Structure"]
        direction TB
        A[QFeatures Container] --> B[colData<br/>Sample Metadata]
        A --> C[Assays<br/>Hierarchical Levels]
        A --> D[rowData<br/>Feature Annotations]
        
        C --> E1[PSMs Assay<br/>Rows: 5000 PSMs<br/>Cols: 10 Samples]
        C --> E2[Peptides Assay<br/>Rows: 2500 Peptides<br/>Cols: 10 Samples]
        C --> E3[Proteins Assay<br/>Rows: 800 Proteins<br/>Cols: 10 Samples]
        
        E1 -->|aggregateFeatures<br/>by Sequence| E2
        E2 -->|aggregateFeatures<br/>by Protein| E3
    end
    
    subgraph Meta["Metadata Propagation"]
        F[Sample Info<br/>Condition, Batch, etc.] --> B
        G1[PSM Annotations<br/>Scores, RT, m/z] --> D
        G2[Peptide Info<br/>Sequence, Modifications] --> D
        G3[Protein Info<br/>Accession, Gene] --> D
    end
    
    subgraph Process["Data Processing"]
        E3 --> H[filterNA<br/>Remove Missing]
        H --> I[normalize<br/>Median/Quantile]
        I --> J[impute<br/>KNN/MinProb]
        J --> K[logTransform<br/>log2]
        K --> L[limma Analysis<br/>Differential Expression]
    end
    
  style QF fill:#D7E6FB,stroke:#27408B,stroke-width:3px,color:#102A43
  style Meta fill:#FBE0FA,stroke:#B000B0,stroke-width:3px,color:#102A43
  style Process fill:#D7E6FB,stroke:#27408B,stroke-width:3px,color:#102A43
```

::: {.callout-note}
## QFeatures Advantages

- **Traceability**: Links between PSMs, peptides, and proteins maintained throughout
- **Flexibility**: Multiple assays can coexist (different processing strategies)
- **Metadata**: Sample and feature annotations travel with the data
- **Reproducibility**: Complete processing pipeline encoded in object
:::

```{r}
# Load example data
data(feat1)
feat1
```

```{r}
# Examine the structure
colData(feat1)
```

```{r}
# Access the PSM-level assay
psms_assay <- feat1[["psms"]]
psms_assay
```

```{r}
# View quantitative data
assay(psms_assay)
```

```{r}
# Examine row annotations
rowData(psms_assay)
```

### Feature Aggregation

A key feature of `QFeatures` is the ability to aggregate features from lower to higher levels while maintaining traceability.

```{r}
# Aggregate PSMs to peptides based on sequence
feat1 <- aggregateFeatures(feat1, 
                          i = "psms",
                          fcol = "Sequence", 
                          name = "peptides",
                          fun = colMeans)
feat1
```

```{r}
# Examine the peptide-level data
assay(feat1[["peptides"]])
```

```{r}
# Check aggregation statistics
rowData(feat1[["peptides"]])
```

```{r}
# Aggregate peptides to proteins
feat1 <- aggregateFeatures(feat1,
                          i = "peptides", 
                          fcol = "Protein",
                          name = "proteins",
                          fun = colMedians)
feat1
```

```{r}
# View final protein quantification
assay(feat1[["proteins"]])
```

### Subsetting and Filtering

`QFeatures` maintains relationships between assays during subsetting operations.

```{r}
# Subset for a specific protein
protein_a <- feat1["ProtA", , ]
protein_a
```

```{r}
# Filter features based on quality criteria
feat1_filtered <- filterFeatures(feat1, ~ pval < 0.05)
feat1_filtered
```

## Working with Real Data: CPTAC Dataset

### Data Import

```{r}
library(MsDataHub)

# Load CPTAC peptide data
# Note: This is a simulated example based on the CPTAC study design
set.seed(123)

# Create sample metadata
sample_info <- data.frame(
  sample = paste0("Sample_", 1:6),
  condition = rep(c("6A", "6B"), each = 3),
  replicate = rep(1:3, 2),
  row.names = paste0("Sample_", 1:6)
)

# Simulate peptide quantification data
n_peptides <- 1000
peptide_data <- matrix(
  rlnorm(n_peptides * 6, meanlog = 10, sdlog = 1),
  nrow = n_peptides,
  ncol = 6,
  dimnames = list(
    paste0("Peptide_", 1:n_peptides),
    rownames(sample_info)
  )
)

# Add differential expression signal
de_peptides <- 1:100  # First 100 peptides are differentially expressed
peptide_data[de_peptides, 4:6] <- peptide_data[de_peptides, 4:6] * 1.5

# Create row annotations
peptide_annotations <- data.frame(
  Sequence = paste0("SEQ", 1:n_peptides),
  Proteins = sample(paste0("PROT", 1:200), n_peptides, replace = TRUE),
  PEP = runif(n_peptides, 0, 0.1),
  Score = runif(n_peptides, 20, 100),
  row.names = rownames(peptide_data)
)

# Create SummarizedExperiment
library(SummarizedExperiment)
cptac_se <- SummarizedExperiment(
  assays = list(peptides = peptide_data),
  rowData = peptide_annotations,
  colData = sample_info
)

# Create QFeatures object
cptac_qf <- QFeatures(list(peptides = cptac_se))
cptac_qf
```

### Data Preprocessing Pipeline

```{r}
# Log transformation
cptac_qf <- logTransform(cptac_qf, 
                        i = "peptides",
                        name = "log_peptides")

# Normalization (median centering)
cptac_qf <- normalize(cptac_qf,
                     i = "log_peptides", 
                     name = "norm_peptides",
                     method = "center.median")

cptac_qf
```

### Missing Value Analysis

```{r}
# Introduce some missing values for demonstration
assay_data <- assay(cptac_qf[["norm_peptides"]])
# Set 10% of values to NA randomly
missing_indices <- sample(length(assay_data), length(assay_data) * 0.1)
assay_data[missing_indices] <- NA
assay(cptac_qf[["norm_peptides"]]) <- assay_data

# Analyze missing value patterns
na_stats <- nNA(cptac_qf[["norm_peptides"]])
cat("Overall missing values:", na_stats$nNA$pNA * 100, "%\n")
```

```{r}
# Visualize missing value patterns
missing_pattern <- na_stats$nNArows
head(missing_pattern, 10)
```

```{r}
# Filter peptides with too many missing values
cptac_qf_clean <- filterNA(cptac_qf, i = "norm_peptides", pNA = 0.5)
cat("Peptides after filtering:", nrow(cptac_qf_clean[["norm_peptides"]]), "\n")
```

### Protein Aggregation

```{r}
# Aggregate to protein level using median
cptac_qf_clean <- aggregateFeatures(cptac_qf_clean,
                                   i = "norm_peptides",
                                   fcol = "Proteins", 
                                   name = "proteins",
                                   fun = colMedians,
                                   na.rm = TRUE)

cptac_qf_clean
```

```{r}
# Examine aggregation results
aggregation_stats <- rowData(cptac_qf_clean[["proteins"]])[".n"]
table(aggregation_stats)
```

## Quality Control and Visualization

### Principal Component Analysis

```{r}
# PCA on peptide-level data
peptide_pca <- cptac_qf_clean[["norm_peptides"]] %>%
  filterNA() %>%
  assay() %>%
  t() %>%
  prcomp(scale = TRUE, center = TRUE)

# Create PCA plot
if (factoextra_available) {
  factoextra::fviz_pca_ind(
    peptide_pca,
    habillage = colData(cptac_qf_clean)$condition,
    title = "Peptide-level PCA"
  )
} else {
  peptide_scores <- as.data.frame(peptide_pca$x[, 1:2])
  peptide_scores$condition <- colData(cptac_qf_clean)$condition

  ggplot(peptide_scores, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3, alpha = 0.8) +
    labs(title = "Peptide-level PCA") +
    theme_minimal()
}
```

```{r}
# PCA on protein-level data  
protein_pca <- cptac_qf_clean[["proteins"]] %>%
  filterNA() %>%
  assay() %>%
  t() %>%
  prcomp(scale = TRUE, center = TRUE)

if (factoextra_available) {
  factoextra::fviz_pca_ind(
    protein_pca,
    habillage = colData(cptac_qf_clean)$condition,
    title = "Protein-level PCA"
  )
} else {
  protein_scores <- as.data.frame(protein_pca$x[, 1:2])
  protein_scores$condition <- colData(cptac_qf_clean)$condition

  ggplot(protein_scores, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3, alpha = 0.8) +
    labs(title = "Protein-level PCA") +
    theme_minimal()
}
```

### Expression Profile Visualization

```{r}
# Extract data for a specific protein using longForm
example_protein <- rownames(assay(cptac_qf_clean[["proteins"]]))[1]

profile_data <- longForm(cptac_qf_clean[example_protein, , 
                                       c("norm_peptides", "proteins")]) %>%
  as_tibble()

# Get column data and ensure unique column names
col_data <- as_tibble(colData(cptac_qf_clean), rownames = "sample_id")

# Join the data
profile_data <- profile_data %>%
  left_join(col_data, by = c("colname" = "sample_id"))

ggplot(profile_data, aes(x = colname, y = value, color = condition)) +
  geom_point(size = 3) +
  geom_line(aes(group = rowname)) +
  facet_wrap(~assay, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = paste("Expression Profile:", example_protein),
       x = "Sample", y = "Normalized Intensity")
```

## Statistical Analysis

### Differential Expression with limma

```{r}
# Extract protein data for statistical analysis
protein_data <- getWithColData(cptac_qf_clean, "proteins")

# Set up design matrix
design <- model.matrix(~ condition, data = colData(protein_data))
colnames(design) <- c("Intercept", "Condition_6B_vs_6A")

# Fit linear model
fit <- lmFit(assay(protein_data), design)
fit <- eBayes(fit)

# Extract results
results <- topTable(fit, coef = "Condition_6B_vs_6A", number = Inf) %>%
  rownames_to_column("protein") %>%
  as_tibble()

head(results)
```

```{r}
# Volcano plot
results %>%
  ggplot(aes(x = logFC, y = -log10(P.Value))) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
  labs(title = "Volcano Plot",
       x = "log2 Fold Change", 
       y = "-log10 P-value") +
  theme_minimal()
```

```{r}
# Summary of differential expression
significant_proteins <- results %>%
  filter(adj.P.Val < 0.05, abs(logFC) > 1)

cat("Significantly changed proteins:", nrow(significant_proteins), "\n")
cat("Up-regulated:", sum(significant_proteins$logFC > 1), "\n")
cat("Down-regulated:", sum(significant_proteins$logFC < -1), "\n")
```

### Heatmap of Significant Proteins

```{r}
if (nrow(significant_proteins) > 5) {
  # Select top 20 most significant proteins
  top_proteins <- head(significant_proteins, 20)$protein
  
  # Create heatmap data
  heatmap_data <- assay(protein_data)[top_proteins, ]
  
  # Sample annotations
  annotation_col <- data.frame(
    Condition = colData(protein_data)$condition,
    row.names = colnames(heatmap_data)
  )
  
  # Generate heatmap
  pheatmap(heatmap_data,
           annotation_col = annotation_col,
           scale = "row",
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           main = "Top Differentially Expressed Proteins")
}
```

## Where MSstats Fits

The workflow above stays within the Bioconductor `QFeatures` ecosystem: `limma` operates on the protein-level matrix produced by `aggregateFeatures()`. An alternative ecosystem, **MSstats**, approaches quantification from the opposite direction -- it models peptide- or transition-level intensities directly, using linear mixed models, and performs protein-level inference as part of the statistical test rather than as a separate aggregation step.

### The MSstats Family

The MSstats ecosystem has grown into a family of specialised packages, each tailored to a specific input type and experimental design:

| Package | Input | Purpose |
|---------|-------|---------|
| **MSstats** | Peptide- or transition-level intensities | Linear mixed models for label-free, DDA, DIA, and SRM; run-level summarisation with `dataProcess()` and differential testing with `groupComparison()` |
| **MSstatsTMT** | Reporter-ion intensities from TMT experiments | TMT-specific workflow: channel-level normalisation, purity correction, and protein-level inference across multiple TMT mixtures |
| **MSstatsPTM** | PTM + global (unmodified) protein abundances | Adjusts post-translational modification quantifications for changes in the parent protein abundance |
| **MSstatsLiP** | Limited proteolysis (LiP) data | Compares protease-accessible peptide profiles for structural proteomics |
| **MSstatsConvert** | Search-engine/vendor export files | Import and conversion layer: transforms output from DIA-NN, FragPipe, MaxQuant, Spectronaut, Skyline, and others into the MSstats-compatible format |
| **MSstatsShiny** | (GUI wrapper) | Point-and-click interface for interactive exploration; generates reproducible R scripts from the GUI session |
| **MSstatsBig** | Large-scale DIA datasets (>10 GB PSM files) | Handles datasets too large for in-memory processing via the `arrow` backend; pre-filters features before normalisation to reduce memory footprint |

### Standard MSstats Workflow

A typical MSstats analysis proceeds through four steps:

1. **Import and convert** --- `MSstatsConvert::MSstatsConvert()` reads search-engine output and converts it to the standard MSstats format
2. **Summarise** --- `MSstats::dataProcess()` performs run-level summarisation, normalisation, and missing-value handling
3. **Compare** --- `MSstats::groupComparison()` fits linear mixed models and returns differential-abundance results
4. **Visualise** --- `MSstats::groupComparisonPlots()` produces volcano plots, comparison plots, and heatmaps

```{r}
#| eval: false
# Typical MSstats pipeline
library(MSstats)

# Convert search-engine output
# raw <- MSstatsConvert::MSstatsConvert(your_export_file)

# Process and summarise
# processed <- dataProcess(raw)

# Compare conditions
# comparison <- groupComparison(processed)

# Visualise
# groupComparisonPlots(comparison, type = "Volcano")
```

### Version Sensitivity of Converters

The `MSstatsConvert` package is **version-sensitive**: different versions of search engines (MaxQuant, DIA-NN, FragPipe, Spectronaut, Skyline) may produce output formats that require specific converter versions. Always check the MSstats news and the converter vignette for the version compatibility table. If a converter fails, the first troubleshooting step is to update both the search engine and `MSstatsConvert` to their latest releases.

### QFeatures vs MSstats: A Design Choice

The `limma` + `QFeatures` workflow demonstrated earlier in this chapter stays within the Bioconductor ecosystem. The MSstats family uses its own data structures and modelling framework. Both approaches are valid; the choice depends on your input format and desired model complexity:

- **Use QFeatures + limma** when you want full control over each preprocessing step, need to integrate with other Bioconductor objects (`SummarizedExperiment`, `MultiAssayExperiment`), or prefer a separation of aggregation and testing.
- **Use MSstats** when your data comes from search engines with native MSstats converters, you need peptide-level modelling with automatic protein-level inference, or you work with complex multi-factorial designs that benefit from the linear mixed model framework.

### Exercises

1. Map three vendor/search-tool exports to the correct converter and downstream MSstats workflow: (a) MaxQuant `proteinGroups.txt`, (b) DIA-NN `report.tsv`, (c) Spectronaut `xyz_Report.tsv`.
2. Compare the output of `MSstats::dataProcess()` with the protein-level matrix produced by `aggregateFeatures()` --- how do the summarised values differ?
3. Explain why `MSstatsConvert` version matters when working with a recently updated version of DIA-NN or FragPipe.

## Advanced Aggregation Methods

### Robust Summarization

```{r}
library(MsCoreUtils)

# Robust aggregation using robust summarization
cptac_robust <- aggregateFeatures(cptac_qf_clean,
                                 i = "norm_peptides",
                                 fcol = "Proteins",
                                 name = "proteins_robust", 
                                 fun = MsCoreUtils::robustSummary,
                                 na.rm = TRUE)

# Compare standard vs robust aggregation
comparison_data <- data.frame(
  standard = assay(cptac_qf_clean[["proteins"]])[, 1],
  robust = assay(cptac_robust[["proteins_robust"]])[, 1]
) %>%
  na.omit()

ggplot(comparison_data, aes(x = standard, y = robust)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Standard vs Robust Aggregation",
       x = "Standard (Median)",
       y = "Robust Summarization") +
  theme_minimal()
```

## Working with QFeatures Workflows

### Visualization of Data Relationships

```{r}
# Visualize the relationship between assays
plot(cptac_robust)
```

### Custom Processing Functions

```{r}
# Example: Custom normalization using addAssay
quantile_normalize <- function(x) {
  # Simple quantile normalization implementation
  x_sorted <- apply(x, 2, sort, na.last = TRUE)
  x_mean <- rowMeans(x_sorted, na.rm = TRUE)
  
  for (i in 1:ncol(x)) {
    ranks <- rank(x[, i], na.last = "keep")
    x[, i] <- x_mean[ranks]
  }
  return(x)
}

# Apply custom normalization
log_peptides_se <- cptac_qf_clean[["log_peptides"]]
quantile_norm_assay <- quantile_normalize(assay(log_peptides_se))

# Create a new SummarizedExperiment with the normalized data
library(SummarizedExperiment)
quantile_norm_se <- SummarizedExperiment(
  assays = list(quantile_norm = quantile_norm_assay),
  rowData = rowData(log_peptides_se),
  colData = colData(log_peptides_se)
)

# Add to QFeatures object
cptac_custom <- addAssay(cptac_qf_clean,
                        quantile_norm_se,
                        name = "quantile_norm")

# Add assay link - for a transformation, use simple string-based linking
# This indicates that all features in quantile_norm come from log_peptides
cptac_custom <- addAssayLink(cptac_custom,
                            from = "log_peptides",
                            to = "quantile_norm")

cat("Custom normalization applied successfully\n")
cat("Available assays:", names(cptac_custom), "\n")
```

## Exercises

1. Load the CPTAC dataset and perform complete preprocessing pipeline
2. Compare different aggregation methods (mean, median, robust)
3. Implement missing value imputation strategies
4. Perform differential expression analysis with multiple comparisons
5. Create custom visualization functions for QFeatures objects

## Summary

This chapter introduced the `QFeatures` framework for quantitative proteomics analysis. Key concepts covered include:

- Different quantitation methodologies in proteomics
- The hierarchical structure of MS quantitative data
- Feature aggregation strategies
- Quality control and missing value handling
- Statistical analysis workflows
- Visualization of quantitative proteomics data

The `QFeatures` infrastructure provides a robust foundation for reproducible quantitative proteomics analysis in R.

## Session Information

``` r
sessionInfo()
```
