# Normalize Across Samples and Batches

> *“Before you ask which proteins or metabolites changed, you need to ask whether the matrix is trustworthy.”*

Load the same pooled sample twice and you will get two different numbers. Run a study across two weeks, two columns, or two reagent lots, and that technical variation can swamp the biology you came to measure — or, worse, mimic it. Normalization and batch correction are how you make samples comparable without scrubbing away the signal. Do too little and batch effects masquerade as discoveries; do too much and you erase the differences that are real.

::: {.callout-warning title="The One Mistake to Avoid"}
Correcting for batch *twice* — once with ComBat and again as a model covariate. Removing the effect twice biases your inference. Correct for visualization, or model the covariate for testing; never both.
:::

## Learning Objectives

By the end of this chapter you will be able to:

- Log‑transform and scale feature matrices correctly
- Apply common normalization methods: median, quantile, LOESS, and TIC‑based
- Diagnose and correct batch effects using ComBat
- Compare normalization methods using CV and PCA
- Produce an analysis‑ready intensity matrix with appropriate QC visualisations

## Why Normalization Is Critical for MS Data

Raw MS feature tables contain systematic non‑biological variation:

| Source | Example | Consequence |
|--------|---------|-------------|
| **Sample loading** | Different protein/metabolite amounts injected | Global intensity shifts |
| **Instrument sensitivity** | Detector aging over time | Drift across injection order |
| **Batch effects** | Different columns, reagent lots, operators | Discontinuous jumps |
| **Run‑to‑run variation** | Mobile phase, temperature, flow rate | Unpredictable changes |

Normalization does **not** remove all unwanted variation – that requires careful experimental design and batch correction. But it is the essential first step to make samples comparable.

### Choosing a Normalization Method

The choice of normalization method depends on the structure of the unwanted variation and the properties of your data. There is no universally "best" method — the choice is a modeling decision, not a default:

| Method | Assumption | Best When | Weakness |
|--------|-----------|-----------|----------|
| **Median centering** | Systematic shift is additive and uniform across features | Small datasets, simple designs | Cannot correct non-linear or feature-dependent bias |
| **Quantile normalisation** | Feature intensity distributions are identical across samples (up to shift) | Microarrays; large $p$ with similar distributions | Distorts distributions when they differ fundamentally |
| **Cyclic LOESS** | Bias is smooth in intensity space | Dye-bias in two-colour arrays; MS when bias varies with intensity | Computationally expensive for many samples |
| **TIC normalisation** | Total ion current is proportional to sample amount | LC-MS with consistent sample composition | Fails when a few highly abundant features dominate TIC |
| **Internal standard (IS)** | IS behaves identically to analytes of interest | Targeted MS with stable-isotope IS | Requires IS; IS may degrade or ionise differently |
| **ComBat** (batch correction) | Batch effects are additive on the log scale; batches share biological composition | Multi-batch studies with balanced designs | Over-corrects if batches are confounded with biology |

### The Decision Rule for MS Data

For untargeted LC-MS metabolomics or proteomics, a robust workflow is:

1.  **Log-transform** first — MS intensity data are log-normal; normalisation on the linear scale amplifies high-abundance features
2.  **Median normalisation** as default — robust to outliers, minimal assumptions
3.  **Assess with PCA** before and after — samples should cluster by biology, not batch
4.  **Apply batch correction** (ComBat) only if batch effects persist after median normalisation AND batches are not confounded with biological groups
5.  **Validate with QC pools** — CV in pooled QC samples should decrease after normalisation

**Never skip PCA before/after normalisation.** If normalisation fails to separate samples by biological group or fails to reduce technical clustering, the downstream analysis will produce artefacts regardless of statistical sophistication.

This chapter assumes you already have a **feature table** (rows = features, columns = samples) from preprocessing (e.g., `xcms` or `DEP`). We will work with a realistic simulated dataset and also show how to adapt to real `msdata` examples.

---

## Setup and Required Packages

```{r}
#| eval: false
BiocManager::install(c(
  "limma",        # Normalization functions
  "preprocessCore", # Quantile normalization
  "impute",       # k‑NN imputation
  "msdata",
  "ggplot2",
  "dplyr",
  "tidyr"
))
```

```{r}
library(limma)
library(preprocessCore)
library(impute)
library(msdata)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(42)
```

---

## Step 1: From Feature Table to Analysis‑Ready Matrix

### Simulate a Realistic MS Feature Table

We simulate 150 metabolites and 24 samples (3 batches, 2 groups). True biological signal exists for the first 30 metabolites.

```{r}
n_metabolites <- 150
n_samples <- 24
n_batches <- 3
samples_per_batch <- n_samples / n_batches

# Sample metadata
metadata <- data.frame(
  sample_id = paste0("S", 1:n_samples),
  batch = rep(paste0("Batch", 1:n_batches), each = samples_per_batch),
  condition = rep(rep(c("Control", "Treatment"), each = samples_per_batch/2), times = n_batches),
  injection_order = 1:n_samples
)

# Base intensities (log‑normal)
intensity_raw <- matrix(rlnorm(n_metabolites * n_samples, meanlog = 10, sdlog = 1.5),
                        nrow = n_metabolites, ncol = n_samples)
rownames(intensity_raw) <- paste0("Metab_", 1:n_metabolites)
colnames(intensity_raw) <- metadata$sample_id

# Add batch effect: Batch2 has +30% intensity, Batch3 has -20%
batch_effect <- c(0, 0.3, -0.2)
for (i in 1:n_batches) {
  idx <- metadata$batch == paste0("Batch", i)
  intensity_raw[, idx] <- intensity_raw[, idx] * (1 + batch_effect[i])
}

# Add treatment effect for first 30 metabolites (2‑fold up in Treatment)
treatment_idx <- metadata$condition == "Treatment"
intensity_raw[1:30, treatment_idx] <- intensity_raw[1:30, treatment_idx] * 2

# Add missing values (20% random, plus 10% MNAR in treatment group)
set.seed(42)
na_random <- matrix(runif(n_metabolites * n_samples) < 0.2, nrow = n_metabolites, ncol = n_samples)
na_mnar <- matrix(FALSE, nrow = n_metabolites, ncol = n_samples)
# MNAR: low‑intensity features in treatment group
low_intensity <- intensity_raw < median(intensity_raw)
na_mnar[1:30, treatment_idx] <- low_intensity[1:30, treatment_idx] & runif(30 * sum(treatment_idx)) < 0.5
intensity_raw[na_random | na_mnar] <- NA

cat(sprintf("Feature table: %d metabolites × %d samples\n", n_metabolites, n_samples))
cat(sprintf("Overall missing rate: %.1f%%\n", 100 * mean(is.na(intensity_raw))))
```

### Log Transformation

MS intensities are heteroscedastic – variance increases with mean. Log2 transformation stabilises variance and makes multiplicative effects additive.

```{r}
# Log2 transform (add 1 to avoid log(0) if zeros present)
intensity_log <- log2(intensity_raw + 1)

# Compare distribution before/after
plot_df <- data.frame(
  raw = intensity_raw[!is.na(intensity_raw)],
  log = intensity_log[!is.na(intensity_log)]
)
plot_df |>
  pivot_longer(everything(), names_to = "transform", values_to = "intensity") |>
  ggplot(aes(x = intensity, fill = transform)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~transform, scales = "free") +
  labs(title = "Raw vs log2‑transformed intensities") +
  theme_minimal()
```

---

## Step 2: Scaling and Centering

After log transformation, we often center and scale to give each feature equal influence in multivariate analysis.

- **Centering**: subtract mean → features have mean zero  
- **Scaling**: divide by standard deviation → features have unit variance  

```{r}
# Center and scale (robust: use median and MAD)
center_scale <- function(mat, robust = TRUE) {
  if (robust) {
    center <- apply(mat, 1, median, na.rm = TRUE)
    scale <- apply(mat, 1, mad, na.rm = TRUE)
    scale[scale == 0] <- 1
  } else {
    center <- rowMeans(mat, na.rm = TRUE)
    scale <- apply(mat, 1, sd, na.rm = TRUE)
    scale[scale == 0] <- 1
  }
  scaled <- (mat - center) / scale
  return(scaled)
}

intensity_scaled <- center_scale(intensity_log, robust = TRUE)

# Check one metabolite
meta_id <- rownames(intensity_scaled)[1]
data.frame(
  original = intensity_log[meta_id, ],
  scaled = intensity_scaled[meta_id, ]
) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = value)) + geom_histogram(bins = 20) + facet_wrap(~name, scales = "free")
```

**When to use scaling?**  
- **PCA, heatmaps, clustering**: yes – prevents high‑abundance features dominating  
- **Univariate testing (t‑test, limma)**: usually not necessary because each feature is tested independently  

---

## Step 3: Normalization Methods

Normalization adjusts for sample‑wide intensity differences. **Always normalise after log transform** unless using TIC‑based (which works on raw scale).

### Method 1: Median Normalization

Assume most features do not change between samples – scale each sample to have the same median intensity.

```{r}
median_norm <- function(mat) {
  sample_medians <- apply(mat, 2, median, na.rm = TRUE)
  global_median <- median(sample_medians, na.rm = TRUE)
  norm_mat <- sweep(mat, 2, sample_medians / global_median, FUN = "/")
  return(norm_mat)
}

intensity_med_norm <- median_norm(intensity_log)
```

### Method 2: TIC (Total Ion Current) Normalization

Scale each sample by its total sum of intensities. Works on raw or log scale (but log then TIC is unusual). Best on raw scale.

```{r}
tic_norm <- function(mat) {
  # mat is raw intensities (not log)
  sample_sums <- colSums(mat, na.rm = TRUE)
  global_mean <- mean(sample_sums, na.rm = TRUE)
  norm_mat <- sweep(mat, 2, sample_sums / global_mean, FUN = "/")
  return(norm_mat)
}

# Apply to raw intensities (with NA handling)
intensity_raw_imp <- intensity_raw
intensity_raw_imp[is.na(intensity_raw_imp)] <- 0  # temporary for TIC only
intensity_tic_norm <- tic_norm(intensity_raw_imp)
# Then log transform
intensity_tic_norm_log <- log2(intensity_tic_norm + 1)
```

### Method 3: Quantile Normalization

Forces all samples to have the same intensity distribution. Assumes that the overall distribution of features is identical across samples – often too strong for MS data, but can be useful for removing batch effects.

```{r}
# Requires complete data – impute first (simple column median)
intensity_imp <- intensity_log
for (j in 1:ncol(intensity_imp)) {
  intensity_imp[is.na(intensity_imp[, j]), j] <- median(intensity_imp[, j], na.rm = TRUE)
}
intensity_quantile <- normalize.quantiles(intensity_imp)
colnames(intensity_quantile) <- colnames(intensity_imp)
rownames(intensity_quantile) <- rownames(intensity_imp)

# Compare original vs quantile
plot_df <- data.frame(
  original = intensity_log[, 1],
  quantile = intensity_quantile[, 1]
)
ggplot(plot_df, aes(x = original, y = quantile)) + geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Quantile normalization effect (Sample 1)")
```

### Method 4: LOESS Normalization (for injection order drift)

If intensity drifts systematically with injection order, LOESS smoothing can correct it.

```{r}
loess_norm <- function(mat, injection_order) {
  # mat: log‑transformed, rows = features, cols = samples
  norm_mat <- mat
  for (i in 1:nrow(mat)) {
    y <- mat[i, ]
    if (sum(!is.na(y)) > 5) {
      loess_fit <- loess(y ~ injection_order, span = 0.75)
      pred <- predict(loess_fit)
      norm_mat[i, ] <- y - pred + mean(y, na.rm = TRUE)
    }
  }
  return(norm_mat)
}

# Apply (requires injection order in metadata)
injection_order <- metadata$injection_order
intensity_loess_norm <- loess_norm(intensity_log, injection_order)
```

### Compare Normalization Methods

```{r}
# Function to compute median CV per sample (after log)
calc_sample_cv <- function(mat) {
  apply(mat, 2, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
}

cv_df <- data.frame(
  raw = calc_sample_cv(intensity_log),
  median = calc_sample_cv(intensity_med_norm),
  quantile = calc_sample_cv(intensity_quantile),
  tic = calc_sample_cv(intensity_tic_norm_log)
)

cv_df |>
  pivot_longer(everything(), names_to = "method", values_to = "cv") |>
  ggplot(aes(x = method, y = cv)) + geom_boxplot() +
  labs(title = "Sample CV by normalization method", y = "CV (coefficient of variation)")
```

**Interpretation:** Lower median CV indicates better removal of sample‑wide variation. Quantile often gives lowest CV but may over‑correct.

---

## Common Pitfalls and Solutions

| Pitfall | Consequence | Solution |
|---------|-------------|----------|
| Normalising before log transform | Multiplicative effects not stabilised | Always log‑transform first (except TIC) |
| Using quantile normalisation on MS data | Removes true biological variation | Use median or LOESS; quantile only for strong batch effects |
| Applying ComBat before splitting train/test | Test leaks into batch model | Correct batches inside cross‑validation or on training data only |
| Scaling before univariate testing | Inflates or deflates effect sizes | Scale only for multivariate methods (PCA, heatmaps) |
| Skipping PCA check after normalization | Residual batch structure invisible | Always plot PCA coloured by batch before proceeding |

---

## Exercises

### Exercise 1: Compare Normalization Methods

Using the simulated dataset, apply median, quantile, TIC, and LOESS normalization. Which gives the lowest variance within batches? Use boxplots of sample CV.

```{r}
# Your code here
```

### Exercise 2: Batch Effect Severity

Simulate a stronger batch effect (larger `batch_shift`). At what point does median normalization fail? Does ComBat still correct adequately? Use PCA before/after to compare.

```{r}
# Your code here
```

### Exercise 3: LOESS Normalization

Replace median normalization with `limma::normalizeCyclicLoess()`. Compare the sample-to-sample correlation matrices before and after normalization using `pheatmap`.

```{r}
# Your code here
```

### Exercise 4: Full Normalization Pipeline

Write a function that takes a raw feature table and metadata and applies log transform → median normalization → ComBat batch correction. Return the corrected matrix and a PCA plot. Run it on the simulated data.

```{r}
# Your code here
```

---

## Summary

### Key Steps in Order

| Step | Purpose | R Code (example) |
|------|---------|------------------|
| Log transform | Stabilise variance | `log2(x + 1)` |
| Normalize | Remove sample‑wide bias | `median_norm()`, `normalize.quantiles()` |
| Check PCA | Detect residual batch structure | `prcomp()` + `ggplot2` |
| Batch correction | Remove technical batch effects | `sva::ComBat()` |
| (Optional) Scale | Equalise feature variance for multivariate | `scale()` with robust options |

### Recommended Default Workflow for MS Data

1. **Log2 transform**
2. **Normalize**: median or TIC (check which reduces CV best)
3. **Inspect PCA** coloured by batch — if clusters visible, proceed to step 4
4. **ComBat** batch correction (parametric for large batches, non‑parametric for small)
5. **Scale only for multivariate analysis** (PCA, heatmaps, clustering)
6. **Missing values** → handled in the next chapter

### Resources

- [limma normalization functions](https://www.rdocumentation.org/packages/limma/versions/3.48.3/topics/normalizeBetweenArrays)
- [preprocessCore quantile normalization](https://bioconductor.org/packages/release/bioc/html/preprocessCore.html)
- [sva/ComBat batch correction](https://bioconductor.org/packages/release/bioc/html/sva.html)  

---

## Session Information

```{r}
sessionInfo()
```