# Modeling Covariates and Repeated Measures for MS Data

> *"A two‑group t‑test is rarely enough. Real MS experiments have batches, paired samples, and repeated measurements. Ignoring these structures inflates false positives."*

Real studies are messy: the same patient sampled before and after treatment, cell lines processed in batches, age and sex riding along with the condition you actually care about. Treat these correlated, structured measurements as if they were independent and you inflate your confidence and manufacture false positives. This chapter models the structure instead of ignoring it — often the difference between a result that replicates and one that does not.

::: {.callout-warning title="The One Mistake to Avoid"}
Treating repeated measures as independent samples. Paired or longitudinal data analyzed as if independent inflates your effective *n* and your false-positive rate. Block on the subject.
:::

## Learning Objectives

By the end of this chapter, you will be able to:

-   Add covariates (batch, sex, age) to a `limma` design matrix for MS feature tables\
-   Fit **paired** and **blocked** models for matched designs (e.g., before/after treatment)\
-   Use `lme4` for repeated‑measures data with random subject effects (time courses)\
-   Decompose variance sources with `variancePartition` to identify dominant factors\
-   Interpret interaction terms in factorial designs (e.g., treatment × time)\
-   Export adjusted differential abundance tables ready for biological interpretation

## Why Modeling Structure Matters for MS Data

MS‑based omics experiments rarely have a simple two‑group design. Real data often include:

| Structure | Example in MS | Ignoring It Causes |
|------------------|-----------------------|-------------------------------|
| **Batch effects** | Samples run on different days or columns | False positives (batch confounded with treatment) |
| **Paired design** | Same patient before/after drug | Reduced power (between‑subject variance not removed) |
| **Repeated measures** | Time course from same subjects | Inflated significance (pseudoreplication) |
| **Confounders** | Age, sex, BMI in clinical cohorts | Biased estimates of treatment effect |
| **Factorial design** | Treatment × time interaction | Missed interaction effects (e.g., treatment only works at late time) |

**The principle:** Your statistical model must match your experimental design. For MS data, the typical workflow after preprocessing (e.g., with `xcms` or `DEP`) results in a **feature table** (rows = metabolites/peptides, columns = samples). This chapter assumes you already have such a table.

### The ICC: Why You Cannot Ignore Correlation

When the same subject is measured multiple times, observations are **correlated**. The intraclass correlation coefficient (ICC) quantifies this:

$$\rho = \frac{\sigma^2_b}{\sigma^2_b + \sigma^2_e}$$

where $\sigma^2_b$ is the between-subject variance and $\sigma^2_e$ is the within-subject (residual) variance. The effective sample size for a study with $T$ repeated measures per subject is:

$$n_{\text{eff}} = \frac{nT}{1 + (T-1)\rho}$$

At $\rho = 0.5$ and $T = 4$, only **57 %** of your nominal sample contributes to inference. This is not a small correction — it means that ignoring repeated measures can inflate your effective sample size by nearly a factor of 2, producing falsely significant results.

### Mixed Models vs. GEE: Two Frameworks for Correlated MS Data

These are the two dominant approaches for correlated data, and they answer fundamentally different questions:

| Aspect | Mixed Models (lme4) | GEE (geepack) |
|--------|---------------------|---------------|
| **Question answered** | "What is the trajectory for a typical individual?" | "What is the average effect across the population?" |
| **Effect type** | Subject-specific | Population-average |
| **Missing data** | Valid under MAR (uses all available data) | Requires MCAR (or WGEE for MAR) |
| **Interpretation** | Conditional on random effects | Marginal — averaged over all subjects |
| **For linear models** | Same estimates as GEE | Same estimates as mixed models |
| **For binary outcomes** | $\beta_{\text{SS}} > \beta_{\text{PA}}$ | Odds ratios are population-averaged |

For MS time-course experiments, mixed models are generally preferred because:
1.  They handle unbalanced data (different numbers of time points per subject) naturally
2.  They are valid under MAR missing data without explicit imputation
3.  The subject-specific interpretation matches the biological question: "How does this metabolite change within an individual over time?"

### Choosing the Covariance Structure

The covariance structure models *how* observations are correlated. The choice affects standard errors but not (for linear models) the fixed-effect estimates:

| Structure | Formula | When to Use |
|-----------|---------|-------------|
| **Compound symmetry** | All pairs equally correlated | Short time series, balanced designs |
| **Random intercept + slope** | Correlation varies with time distance | Growth trajectories (default choice) |
| **AR(1)** | Correlation decays exponentially with time separation | Equally spaced time points |
| **Unstructured** | Each pair has its own correlation | Small T, no prior structure assumed |

**When in doubt, use random intercept + slope.** It captures the most important correlation pattern — that subjects differ in both baseline level and rate of change — with only 2 extra parameters.

------------------------------------------------------------------------

## Datasets for This Chapter

| Dataset                  | Source       | Best for                            |
|------------------------|----------------------|---------------------------|
| `msdata::metabolomics()` | `msdata`     | Real LC‑MS metabolomics (8 samples) |
| Simulated (included)     | this chapter | Learning concepts with known truth  |
| `sacurine`               | `ropls`      | Covariate adjustment (sex, age)     |

We will start with a **simulated feature table** so you can see exactly how each model recovers true biological signals.

------------------------------------------------------------------------

## Setup and Package Installation

```{r}
#| eval: false
BiocManager::install(c(
  "limma",               # Linear models for MS data
  "lme4",                # Mixed‑effects models
  "variancePartition",   # Variance decomposition
  "broom.mixed",         # Tidy mixed model summaries
  "msdata",              # Example MS data
  "ggplot2",
  "dplyr"
))
```

```{r}
#| eval: false
# Load libraries
library(limma)
library(lme4)
library(variancePartition)
library(ggplot2)
library(dplyr)
library(tidyr)
library(broom.mixed)
library(msdata)   # for real MS files

# Set seed for reproducibility
set.seed(42)
```

------------------------------------------------------------------------

## Step 1: Model Formula Reference Table

Before fitting, identify your experimental structure and choose the corresponding formula.

```{r}
#| eval: false
model_reference <- data.frame(
  Design = c(
    "Two‑group (simple)",
    "Covariate‑adjusted",
    "Paired / Blocked",
    "Two‑way factorial",
    "Time course (fixed effects)",
    "Repeated measures (mixed)"
  ),
  Formula = c(
    "~ condition",
    "~ condition + batch + sex",
    "~ condition + subject_id",
    "~ condition * time",
    "~ condition + time",
    "~ condition + (1 | subject)"
  ),
  Function = c(
    "lmFit() + eBayes()",
    "lmFit() + eBayes()",
    "lmFit() + eBayes()",
    "lmFit() + eBayes()",
    "lmFit() + eBayes()",
    "lmer()"
  )
)

knitr::kable(model_reference, caption = "Model formulas for common MS experimental designs")
```

**Usage:** Match your design to a row, adapt the formula to your column names.

------------------------------------------------------------------------

## Step 2: Simulate a Realistic MS Feature Table

We'll create a feature table with **200 metabolites** and **18 samples**.\
True effects: treatment upregulates first 30 metabolites; batch 2 has a systematic shift; sex has a small effect.

```{r}
#| eval: false
n_metabolites <- 200
n_samples <- 18

# Experimental factors
condition <- factor(rep(c("Control", "Treatment"), each = 9))
batch <- factor(rep(c("B1", "B2", "B3"), times = 6))
sex <- factor(sample(c("M", "F"), n_samples, replace = TRUE))

# Feature table: rows = metabolites, columns = samples
intensity_mat <- matrix(rnorm(n_metabolites * n_samples, mean = 10, sd = 1),
                        nrow = n_metabolites, ncol = n_samples)

# Add batch effect (batch 2 is +1.5)
intensity_mat[, batch == "B2"] <- intensity_mat[, batch == "B2"] + 1.5

# Add sex effect (males +0.2)
intensity_mat[, sex == "M"] <- intensity_mat[, sex == "M"] + 0.2

# Add treatment effect for first 30 metabolites
intensity_mat[1:30, condition == "Treatment"] <- 
  intensity_mat[1:30, condition == "Treatment"] + 2

rownames(intensity_mat) <- paste0("Metab_", 1:n_metabolites)
colnames(intensity_mat) <- paste0("Sample_", 1:n_samples)

# Metadata (samples as rows)
metadata <- data.frame(
  sample = colnames(intensity_mat),
  condition = condition,
  batch = batch,
  sex = sex
)

cat(sprintf("Simulated MS feature table: %d metabolites × %d samples\n",
            n_metabolites, n_samples))
head(metadata)
```

> **Note:** In real MS data, you would replace `intensity_mat` with the output from `featureValues()` from `xcms` or a normalized expression matrix from `DEP`.

------------------------------------------------------------------------

## Step 3: Covariate‑Adjusted Model with `limma` (Batch, Sex, etc.)

### The Problem

If batch differs between control and treatment groups, an unadjusted model will find **batch differences** masquerading as **treatment effects**.

### Build Design Matrix

```{r}
#| eval: false
design_adj <- model.matrix(~ condition + batch + sex, data = metadata)
colnames(design_adj)
```

### Fit the Model for All Metabolites

```{r}
#| eval: false
fit_adj <- lmFit(intensity_mat, design_adj)
fit_adj <- eBayes(fit_adj)

# Extract treatment effect (coefficient "conditionTreatment")
results_adj <- topTable(fit_adj, coef = "conditionTreatment",
                        number = Inf, sort.by = "P")
head(results_adj[, c("logFC", "AveExpr", "P.Value", "adj.P.Val")])
```

### Compare Adjusted vs Unadjusted (Ignoring Batch/Sex)

```{r}
#| eval: false
design_unadj <- model.matrix(~ condition, data = metadata)
fit_unadj <- lmFit(intensity_mat, design_unadj) |> eBayes()
results_unadj <- topTable(fit_unadj, coef = "conditionTreatment", number = Inf)

comparison_df <- data.frame(
  Metabolite = rownames(results_adj)[1:30],
  Adjusted_P = results_adj$P.Value[1:30],
  Unadjusted_P = results_unadj$P.Value[1:30]
)
comparison_df |>
  mutate(ratio = Unadjusted_P / Adjusted_P) |>
  head(10) |>
  knitr::kable(digits = 4)
```

**Expected finding:** Adjusted P‑values are often lower (more significant) because batch variance is removed. In MS data, failing to adjust for batch can completely reverse conclusions.

------------------------------------------------------------------------

## Step 4: Paired / Blocked Model for Matched Designs

### When to Use in MS

-   **Before/after** treatment in same patients (paired samples)\
-   **Technical replicates** from the same biological sample (blocked)\
-   **Matched case‑control** (age/sex matched pairs)

### Simulate Paired MS Data

```{r}
#| eval: false
n_patients <- 9
patient_id <- factor(rep(paste0("Patient_", 1:n_patients), times = 2))
condition_paired <- factor(rep(c("Before", "After"), each = n_patients))

# Feature table: each patient contributes two samples
mat_paired <- matrix(nrow = n_metabolites, ncol = 18)
for (i in 1:n_metabolites) {
  baseline <- rnorm(n_patients, mean = 10, sd = 1)
  effect <- ifelse(i <= 30, 1, 0)   # true after effect for first 30
  mat_paired[i, ] <- c(baseline, baseline + effect + rnorm(n_patients, 0, 0.5))
}
colnames(mat_paired) <- paste0(patient_id, "_", condition_paired)
rownames(mat_paired) <- paste0("Metab_", 1:n_metabolites)

paired_metadata <- data.frame(
  sample = colnames(mat_paired),
  patient = patient_id,
  condition = condition_paired
)
```

### Paired Design Matrix

```{r}
#| eval: false
# Include patient ID to absorb subject‑to‑subject variation
design_paired <- model.matrix(~ condition + patient, data = paired_metadata)
fit_paired <- lmFit(mat_paired, design_paired) |> eBayes()
results_paired <- topTable(fit_paired, coef = "conditionAfter",
                           number = Inf, sort.by = "P")

cat("Significant metabolites (adj.P.Val < 0.05):",
    sum(results_paired$adj.P.Val < 0.05), "\n")
cat("True positives detected:",
    sum(results_paired$adj.P.Val[1:30] < 0.05), "/ 30\n")
```

**Key insight:** Including `patient` as a fixed effect removes between‑patient variability, increasing power to detect the treatment effect. This is the linear model equivalent of a paired t‑test.

------------------------------------------------------------------------

## Step 5: Mixed‑Effects Models for Repeated Measures (Longitudinal MS)

### When to Use

-   **Time course** – same subjects measured at 3+ time points\
-   **Multi‑level designs** – measurements nested within subjects (e.g., cells within patients)

### Simulate Longitudinal MS Data

```{r}
#| eval: false
n_subjects <- 20
n_timepoints <- 4

long_df <- expand.grid(
  subject = paste0("S", 1:n_subjects),
  time = 1:n_timepoints,
  condition = c("Control", "Treatment")
) |>
  mutate(
    subject_effect = rep(rnorm(n_subjects, 0, 0.5), each = n_timepoints * 2),
    treatment_effect = ifelse(condition == "Treatment", 0.3 * time, 0),
    abundance = 10 + subject_effect + treatment_effect + rnorm(n(), 0, 0.3)
  )
head(long_df)
```

### Fit Mixed Model for One Metabolite

```{r}
#| eval: false
fit_lmer <- lmer(abundance ~ condition * time + (1 | subject),
                 data = long_df)
tidy(fit_lmer, effects = "fixed") |> knitr::kable(digits = 3)
```

**Coefficient interpretation for MS time courses:**

| Coefficient | Meaning |
|------------------------------------------|------------------------------|
| (Intercept) | Baseline abundance for Control at time 0 |
| conditionTreatment | Difference between Treatment and Control at time 0 |
| time | Change over time in Control group (slope) |
| conditionTreatment:time | **Interaction** – does treatment change the slope over time? |

### Scaling to Many Metabolites

For real MS data, you can loop over metabolites:

```{r}
#| eval: false
# Function to extract interaction p‑value
get_interaction_p <- function(metabolite_data) {
  # metabolite_data must have columns: abundance, condition, time, subject
  fit <- lmer(abundance ~ condition * time + (1 | subject),
              data = metabolite_data)
  tidy(fit, effects = "fixed") |>
    filter(term == "conditionTreatment:time") |>
    pull(p.value)
}

# Example for one metabolite
p_val <- get_interaction_p(long_df)
cat("Interaction p‑value:", p_val, "\n")
```

> **Note:** For many metabolites, consider using `limma` with `duplicateCorrelation` (faster). Mixed models are more flexible but computationally heavier.

------------------------------------------------------------------------

## Step 6: Variance Partitioning – Which Factors Drive MS Variation?

### Why Do This for MS Data?

Understanding variance sources helps you decide whether to include factors in your model and whether your experiment is well‑designed.

```{r}
#| eval: false
# Prepare data: samples as rows, metabolites as columns
intensity_t <- t(intensity_mat)
colnames(intensity_t) <- paste0("Metab_", 1:n_metabolites)

# Formula for variance components
vp_form <- ~ condition + batch + sex

# Fit variance partition
vp <- fitExtractVarPartModel(intensity_t, vp_form, metadata)

# Plot across all metabolites
plotVarPart(vp, main = "Variance decomposition in MS feature table")
```

**Interpreting the plot for MS data:**

| Component | If high (\>30%) | Implication |
|--------------------|-----------------------------|------------------------|
| **condition** | Good – biological signal is strong | Your model will detect differences |
| **batch** | Problematic – technical artifact dominates | Need batch correction (including batch in model) |
| **sex** | Interesting – sex‑specific effects exist | Include sex as covariate or stratify |
| **Residual** | \>50% – high unexplained noise | Need more replicates or better MS measurement |

------------------------------------------------------------------------

## Step 7: Interaction Models (Factorial Designs) for MS

### When to Test Interactions

Interaction answers: *Does the effect of treatment depend on time (or another factor)?*\
Example: A drug reduces metabolite abundance only after 24 hours.

### Simulate Interaction Data (2×2 design)

```{r}
#| eval: false
time2 <- factor(rep(c("T0", "T24"), each = n_samples / 2))
condition2 <- condition

# Add interaction effect for first 30 metabolites
mat_interact <- intensity_mat
mat_interact[1:30, condition2 == "Treatment" & time2 == "T24"] <-
  mat_interact[1:30, condition2 == "Treatment" & time2 == "T24"] + 1.5

interact_metadata <- data.frame(
  sample = colnames(mat_interact),
  condition = condition2,
  time = time2
)
```

### Fit Interaction Model

```{r}
#| eval: false
design_interact <- model.matrix(~ condition * time, data = interact_metadata)
fit_interact <- lmFit(mat_interact, design_interact) |> eBayes()

# Interaction coefficient (conditionTreatment:timeT24)
results_interact <- topTable(fit_interact, coef = "conditionTreatment:timeT24",
                              number = Inf)
head(results_interact[, c("logFC", "P.Value", "adj.P.Val")])
```

### Visualise the Top Interaction

```{r}
#| eval: false
top_metab <- rownames(results_interact)[1]
plot_df <- data.frame(
  intensity = mat_interact[top_metab, ],
  condition = condition2,
  time = time2
)

ggplot(plot_df, aes(x = time, y = intensity, color = condition, group = condition)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  labs(
    title = paste("Interaction plot for", top_metab),
    subtitle = "Non‑parallel lines indicate a significant interaction",
    x = "Time point", y = "log2 intensity"
  ) +
  theme_minimal()
```

**Interpretation for MS data:** If the lines cross or diverge, the treatment effect changes over time – that's the interaction you are testing.

------------------------------------------------------------------------

## Step 8: Model Selection Decision Flowchart for MS Experiments

``` mermaid
flowchart TD
    A[Start: Your MS feature table & metadata] --> B{Repeated measures on<br>same subjects?}
    
    B -->|Yes – same subject measured multiple times| C{How many time points?}
    C -->|2 time points| D[limma paired model<br>~ condition + subject]
    C -->|3+ time points| E[lmer mixed model<br>~ condition * time + (1|subject)]
    
    B -->|No – independent samples| F{Any known confounders<br>batch, sex, age?}
    F -->|Yes| G[limma covariate‑adjusted<br>~ condition + batch + sex]
    F -->|No| H{More than one factor?}
    
    H -->|Yes – factorial| I[limma interaction model<br>~ factor1 * factor2]
    H -->|No – simple| J[limma two‑group<br>~ condition]
    
    D --> K[Extract coefficient of interest]
    E --> K
    G --> K
    I --> K
    J --> K
    
    K --> L[Apply eBayes shrinkage]
    L --> M[FDR correction (BH)]
    M --> N[Export results table]
```

------------------------------------------------------------------------

## Step 9: Complete Workflow (Real MS Data Ready)

Replace `your_intensity_matrix` with your actual feature table (samples as columns).

```{r}
#| eval: false
# Complete covariate modeling pipeline for MS data
library(limma)
library(ggplot2)
library(dplyr)

# 1. Prepare metadata (must match column order of intensity matrix)
metadata <- data.frame(
  condition = factor(rep(c("Ctrl","Treat"), each = 9)),
  batch     = factor(rep(c("B1","B2","B3"), times = 6)),
  sex       = factor(sample(c("M","F"), 18, replace = TRUE))
)

# 2. Build design matrix (include all known confounders)
design <- model.matrix(~ condition + batch + sex, data = metadata)

# 3. Fit model
fit <- lmFit(your_intensity_matrix, design) |> eBayes()

# 4. Extract results for condition effect
results <- topTable(fit, coef = "conditionTreatment",
                    number = Inf, sort.by = "P")

# 5. Filter significant features (FDR < 0.05)
sig_results <- results |> filter(adj.P.Val < 0.05)

# 6. Export
write.csv(sig_results, "MS_differential_results_covariate_adjusted.csv")

# 7. Quick variance partition diagnostic (samples as rows)
library(variancePartition)
vp <- fitExtractVarPartModel(t(your_intensity_matrix),
                              ~ condition + batch + sex,
                              metadata)
plotVarPart(vp)
```

------------------------------------------------------------------------

## Common Pitfalls in MS Data Modeling

| Pitfall | Why It's a Problem | Solution |
|------------------|-----------------------------------|-------------------|
| **Adding too many covariates** | Overfitting, reduced degrees of freedom | Include only known confounders (batch, sex, age) – not all possible variables |
| **Ignoring batch when it's confounded** | Batch effect masquerades as treatment effect | Always check batch association with condition (e.g., χ² test) |
| **Using mixed models for \<5 subjects** | Random effects poorly estimated | Use paired design with fixed subject effect instead |
| **Forgetting interactions** | Main effect may not tell full story (e.g., treatment only works at late time) | Plot data; test interaction if biologically plausible |
| **Treating time as categorical with many levels** | Loss of power, multiple testing burden | Consider continuous time with polynomial or spline terms |
| **Not checking model assumptions** | Residuals may not be normal, heteroscedasticity | Use `plot(fit)` for limma; `plot(fit_lmer)` for mixed models |
| **Matrix orientation confusion** | `lmFit` expects rows = features, columns = samples | Verify `dim(your_intensity_matrix)` – transpose if needed |

------------------------------------------------------------------------

## Exercises (Apply to Your Own MS Data)

### Exercise 1: Add a Continuous Covariate

Using the simulated data from Step 2, add `age` as a continuous covariate (simulate random ages 20‑80). Fit a model with `~ condition + batch + sex + age`. How do the logFC and p‑values for the treatment effect change compared to the model without age?

```{r}
#| eval: false
# Your code here
```

### Exercise 2: Compare Paired vs Unpaired on Real‑World MS Data

Load the `msdata::metabolomics()` files, process them with `xcms` (Chapter 7), and extract a feature table. If your experiment had paired samples (e.g., same subject before/after), run both a paired and an unpaired `limma` model. How many features are significant only in the paired model?

```{r}
#| eval: false
# Your code here
```

### Exercise 3: Variance Partition on a Clinical MS Dataset

Load the `sacurine` dataset from `ropls`. Use `variancePartition` to quantify variance explained by `sample_type` (urine), `age`, and `gender`. Which factor dominates?

```{r}
#| eval: false
# Your code here
```

### Exercise 4: Discover Interactions in a Time‑Course MS Experiment

Using the interaction‑simulated data, test all metabolites for `condition × time` interaction. Report: - How many true positives (first 30 metabolites) are significant at FDR \< 0.05? - How many false positives (metabolites 31‑200) are significant?

```{r}
#| eval: false
# Your code here
```

------------------------------------------------------------------------

## Summary

### Key Outputs from This Chapter (for MS Data)

| Output | Purpose |
|----------------------------------|--------------------------------------|
| Covariate‑adjusted results | Differential abundance controlling for batch, sex, age |
| Paired model results | Increased power for matched designs (before/after) |
| Mixed model summary | Fixed (treatment, time) and random (subject) effects |
| Variance partition plot | Which factors (condition, batch, sex) drive MS variation? |
| Interaction plot | Visualise condition × time effects |
| Model decision table | Choose the right formula for your MS experiment design |

### Model Selection Quick Reference for MS Experiments

| Your MS Design | Use This Formula |
|----------------------------------|--------------------------------------|
| 2 groups, no confounders | `~ condition` |
| 2 groups + batch/sex/age | `~ condition + covariate` |
| Matched pairs (before/after) | `~ condition + subject_id` |
| Time course, 2 time points | `~ condition * time` (interaction) |
| Time course, 3+ time points | `lmer(value ~ condition * time + (1\|subject))` |
| Multiple variance sources unknown | `variancePartition` first, then model |

### Resources Specific to MS Data

-   [limma for proteomics](https://www.bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/limmaUsersGuide.pdf) (Chapter 20)
-   [MSstats mixed models](https://www.bioconductor.org/packages/release/bioc/html/MSstats.html) – designed for MS repeated measures
-   [variancePartition paper with MS example](https://doi.org/10.1186/s12859-016-1323-7)

------------------------------------------------------------------------

## Session Information

```{r}
#| eval: false
sessionInfo()
```
